---
title: "caretSDM"
author: "Luíz Fernando Esser"
format: html
   code-link: true
editor: visual
---

# caretSDM

caretSDM is a under development R package that uses the powerful caret package as the main engine to obtain Species Distribution Models. As caret is a packaged turned to build machine learning models, caretSDM has a strong focus on this approach.

## Installing

First we will install the package from github. For that we will need to install the `devtools` package first and then install the caretSDM package.

```{r}
install.packages(setdiff("devtools", rownames(installed.packages())))
devtools::install_github('luizesser/caretSDM')
```

## Obtaining example data

Now we will create some example data. Starting with occurrences and then predictors data. The first one is a function to obtain species data from GBIF, while the second is a function to obtain climatic variables from WorldClim 2.1. You can read more about them by running in the console `?GBIF_data` and `?WorldClim_data`.

```{r}
occ_data <- GBIF_data("Araucaria angustifolia")
pred_data <- WorldClim_data()
```

## The occurrences class

In this package we will create some objects with specific classes. This objects will be used in different functions.

```{r}
occ <- occurrences(occ_data)
occ
```

```{r}
pred <- predictors(pred_data)
```

## 
