---
title: "Araucaria"
author: "Luíz Fernando Esser"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Araucaria}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Using polygons as area in caretSDM

`caretSDM` is a under development R package that uses the powerful `caret` package as the main engine to obtain Species Distribution Models. One of its main attributes is the strong geoprocessing underlying its functions. Here we show how to model species distributions using `caretSDM` through the function `sdm_area` with a polygon.

```{r open libraries}
library(chooseGCM)
library(caretSDM)
library(stars)
library(dplyr)
```

## Pre-Processing

### Obtaining example data

We will create some example data. Starting with occurrences and then predictors data. The first one is a function to obtain species data from GBIF, while the second is a function to obtain climatic variables from WorldClim 2.1. You can read more about them by running in the console `?GBIF_data` and `?WorldClim_data`.

```{r GBIF_data}
# Import data from GBIF
occ_data <- GBIF_data('Araucaria angustifolia')

# Remove records from outside the natural occurrence area
occ_data <- filter(occ_data, decimalLongitude >= -64)
occ_data <- filter(occ_data, decimalLongitude <= -36)
occ_data <- filter(occ_data, decimalLatitude <= -10)
occ_data <- filter(occ_data, decimalLatitude >= -80)
```

```{r worldclim_data}
## Download current bioclimatic variables
WorldClim_data(period = "current", resolution = 10)

## Enter bioclimatic variables
pred_data <- read_stars(list.files("input_data/WorldClim_data_current/", full.names = T), along = "band", normalize_path = F)
```

```{r study_area}
study_area <- geobr::read_state("PR")
```

```{r create_buffer}
buf <- st_as_sf(occ_data, coords=c(2,3))
buf <- st_buffer(buf, dist=5)
buf <- st_union(buf)
st_crs(buf) <- 4326
```

```{r sdm_area}
sa <- sdm_area(study_area, cell_size = 25000, crs = 6933)
```

```{r add_predictors}
predictors(sa)
sa <- add_predictors(sa, pred_data)
predictors(sa)
```

```{r input_sdm}
i_sa <- input_sdm(occurrences_sdm(occ_data, epsg = 4326), sa)
```

```{r data_clean}
# funciona
i_sa <- data_clean(i_sa)
```

```{r vif_predictors}
i_sa <- vif_predictors(i_sa)
```

```{r}
# Follow with the same SDM workflow:
i_sa <- input_sdm(occurrences_sdm(occ_data, epsg = 4326), sa) |>
  data_clean() |>
  vif_predictors() |>
  pseudoabsences(method = "bioclim", variables_selected = c("bio1", "bio12", "LENGTH_KM")) |>
  train_sdm(algo = c("svmLinear2", "mda", "nnet", "nb", "kknn"), variables_selected = c("bio1", "bio12", "LENGTH_KM")) |>
  predict_sdm(th = 0.9)

```

```{r}
i <- vif_predictors(predictors(s,
                               study_area = buf),
                    area='all',
                    variables_selected=c("bio1","bio2","bio3","bio4","bio5",
                                         "bio6","bio7","bio8","bio9","bio10",
                                         "bio11","bio12","bio13","bio14","bio15",
                                         "bio16","bio17"))


## Get VIF variable selection
var_names <- predictors_names(i)

# Run chooseGCM
## Import GCMs data
s_fut <- import_gcms('/Users/luizesser/Documents/GitHub/input_data/WorldClim_data_future_all') # Folder with all GCMs to use in choose GCMs
names(s_fut) <- gsub("_ssp585_10_2090","",names(s_fut))

## generate plots
var_names2 <- c('bio10', 'bio13', 'bio14', 'bio02')
res <- compare_gcms(s_fut, var_names2, study_area, k = 3)
res

## Download future data given chooseGCMs result
#chooseGCM::WorldClim_data(path = 'input_data/WorldClim_data_future', period = 'future', variable = 'bioc', year = c('2050','2090'), gcm = as.character(res$suggested_gcms), ssp = c('245','585'), resolution = 10)

# Run SDM
## Put all data together
i <- input_sdm(occurrences(occ_data),
               predictors(s,
                          study_area = buf),
               scenarios("input_data/WorldClim_data_future",
                         study_area = study_area))

## Data cleaning routine
i <- data_clean(i)

## Pseudoabsence selection
i <- pseudoabsences(i, method='bioclim', variables_selected=var_names)

## Modelling
library(caret)
ctrl <- trainControl(method = "repeatedcv", number = 4,
                     repeats = 10, classProbs = TRUE, returnResamp = "all",
                     summaryFunction = twoClassSummary, savePredictions = "all")
i <- train_sdm(i,  algo=c('svmLinear2','mda','nnet','nb', 'kknn'), ctrl=ctrl, variables_selected=var_names)

## Predictions and ensembling
i <- predict_sdm(i, th=0.9)

## Plot ensembles:
plot_predictions(i, scenario = 'uk_ssp585_10_2090')

# Export data
## Export rasters
write_ensembles(i)

## Obtain important data
mean_validation_metrics(i)

## Export i
saveRDS(i, 'Araucaria_angustifolia.Rds')
##### End of Script #####


x4 <- Sys.time()
```
