---
title: "Multispecies Distribution Modeling with caretSDM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multispecies Distribution Modeling with caretSDM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 80
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(caretSDM)
```

## Introduction

`caretSDM` is a under development R package that uses the powerful `caret`
package as the main engine to obtain Species Distribution Models. One of its
main attributes is the strong geoprocessing underlying its functions. Here we
show how to model species distributions using `caretSDM` through the function
`sdm_area` with a polygon. We will also show how to apply a PCA in predictors
and scenarios to avoid multicolinearity. The aim of this modeling will be to
obtain the distribution of *Araucaria angustifolia*, *Ilex paraguariensis* and
*Myrceugenia euosma*, three co-occurring tree species.

First, we need to open our libraries.

```{r libraries}
library(caretSDM)
library(stars) # Package used to manipulate GIS data
library(dplyr) # Package to improve data handling
```

## Pre-Processing

To obtain models, we will need climatic data and species records. To easily
obtain these data, we have two functions: `WorldClim_data` function downloads
climatic variables from WorldClim 2.1, a widely used open-source database; in
the same way, `GBIF_data` function downloads species records from GBIF, also a
widely used open-source database. You can read more about them by running in the
console `?GBIF_data` and `?WorldClim_data`.

### Obtaining species records

A easy way to get species data using `caretSDM` is the function `GBIF_data`,
which retrieves species records from GBIF. Understandably, there are other
sources of species data available, as well as our own data that can be the
result of field work. In this sense, one can import to R it's own data in
multiple ways, but be sure that the table will have three columns: species,
decimalLongitude and decimalLatitude. `GBIF_data` function can retrieve the data
ready to be included in caretSDM, thus if you have any doubt on how to format
your own data, use `GBIF_data` function with the parameter `as_df = TRUE` to
retrieve an example table.

```{r}
oc <- GBIF_data(c("Araucaria angustifolia", "Ilex paraguariensis", "Myrceugenia euosma"))
```

### Obtaining climatic data

We will first download and import current data, which is used to build the
models. `WorldClim_data` function has an argument to set the directory in which
you want to save the files. If you don't set it, files will be saved in your
working directory (run `getwd()` to find out your working directory) in the
folder "input_data/WorldClim_data_current/". If period is set to "future", then
it is saved in "input_data/WorldClim_data_future/". We could run this script
with a smaller resolution, but as the aim here is to show how the package works,
we will use a resolution of 10 arc-minutes, which is very coarse, but quicker to
download and run.

```{r WorldClim_data_current}
# Download current bioclimatic variables
WorldClim_data(period = "current", resolution = 10)

# Import current bioclimatic variables to R
pred <- read_stars(list.files("input_data/WorldClim_data_current/", full.names = T), along = "band", normalize_path = F)

# See what is inside pred
pred
```

### Defining the study area

A important step on model building in Species Distribution Models, is the
definition of accessible area (the M in BAM diagram). This area can be, in
Geographical Information Systems terms, as an example, the delimitation of a
habitat (polygon) or a river basin network (lines). Another broadly used
approach is the use of buffers around presences. The buffer size translates the
potential distribution capabilities of a species. To educational purposes we
will use a simple shape with Parana state boundaries that is available in
caretSDM in the `parana` object.

In `caretSDM` there is a function that groups all the transformations regarding
the study area to build and project the models. The `sdm_area` function is this
function and is also responsible to check CRSs and create a grid to build
models. This grid may seem unpurposed for terrestrial ecologists, but is a key
element when modeling continental aquatic environments. `sdm_area` class will
also keep the environmental/climatic data (*i.e.* "predictor variables",
"covariates", "explanatory variables", "features" or "control variables"). With
this class we will perform analysis using only the predictors (such as
predictors transformations).

```{r}
# Create a sdm_area object
sa <- sdm_area(parana, cell_size = 20000, crs = 6933)
sa
plot_grid(sa)
```

Now that we have a study area, we can assign predictor variables and scenarios
to it.

```{r add_predictors1}
# Add predictor variables into sdm_area
sa <- add_predictors(sa, pred)
sa
```

```{r add_scenarios}
sa <- add_scenarios(sa)
sa
```

Usually, we will need to subset variables that will inform our model. This can
be due to statistical artifacts that are common in quarter bioclimatic
variables, or a causation subset, aiming for those variables with causality
effect on species distribution.

```{r select_preds}
# Select predictors
sa <- select_predictors(sa, paste0("X", 1:19, ".tif"))
sa <- set_predictor_names(sa, paste0("bio",1:19))
sa
```

```{r}
write_grid(sa, path = "results/grid_study_area.csv", centroid = TRUE)
write_grid(sa, path = "results/grid_study_area.gpkg", centroid = TRUE)
write_grid(sa, path = "results/grid_study_area.shp", centroid = TRUE)
```

```{r add_predictors}
mapview_predictors(sa)
```

### Defining the occurrences set in the study area

As `caretSDM` has a strong GIS background, it is necessary to explicitly tell
which CRS is your data in. This will assure that every GIS transformation is
correct. This step also assigns occurrences into a study area, excluding records
outside the study area or with NAs in predictors. `occurrences_sdm` function
creates a occurrences class (*i.e.* "response variable", "target" or "label")
that will be used in occurrences transformations and functions, as
pseudoabsences generation. As we used GBIF_data, which

```{r occurrences_sdm}
# Join occurrences to the study area
oc <- oc |>
  join_area(sa)
oc
```

### The `input_sdm` class

In `caretSDM` we use multiple classes to perform our analysis. Every time we
perform a new analysis, objects keep the information of what we did. Ideally,
the workflow will have only one object throughout it. The `input_sdm` class is
the key class in the workflow, where every function will orbitate. That class
puts occurrences, predictors, scenarios, models and predictions together to
perform analysis that are only possible when two or more of these classes are
available. First, we create the object by informing the occurrences and
sdm_area.

```{r input_sdm}
i <- input_sdm(oc, sa)
i
```

```{r}
write_occurrences(i, path = "results/occurrences.csv", grid = FALSE)
write_occurrences(i, path = "results/occurrences.gpkg", grid = TRUE)
write_occurrences(i, path = "results/occurrences.shp", grid = TRUE)
```

Now we can visualize `oc` data using plot:

```{r plot_occurrences}
plot_occurrences(i)
```

### Data cleaning routine

As the first step in our workflow with the `input_sdm` object, we will clean our
occurrences data by applying a group of functions from the package
`CoordinateCleaner`. In this function, we also provide a way to check for
environmental duplicates, by including a predictors object. This function also
checks for records in the sea if the species is terrestrial, but note that this
can be switched off if the studied species is not terrestrial. The way
`caretSDM` works, we can always overwrite the main `input_sdm` object to update
it. The function will return a new object with all the previous information and
the new information obtained from the `data_clean` function, note that at the
end of the Data Cleaning information there is the Duplicated Cell method. This
method is only possible when we have both the `occurrence` and `predictors`
data.

```{r data_clean}
i <- data_clean(i, capitals = FALSE)
i
```

### Removing multicolinearity from predictors' data

In `vif_predictors`, we are able to perform two methods to select variables:
selecting all area or using the presence records (which is debatable, but
implemented). `vif_predictors` allows for a maximum threshold to be informed.
The standard is 0.5.

```{r vif_predictors}
i <- vif_predictors(i, th = 0.5, 
                    variables_selected = c("bio1", "bio2", "bio3", "bio4", "bio5", "bio6", "bio7", "bio12", "bio13", "bio14", "bio15")) 
i
```

### Obtaining pseudoabsence data

As we mentioned before, pseudoabsence data will be stored in the `occurrences`
object (inside the `input_sdm`). To generate them, you can inform some
parameters. The argument `variables_selected` will inform which variables you
want to use to build your pseudoabsences/models. This can either be a vector of
variables names or a previously performed selection method.

```{r pseudoabsence}
#### selecionar variáveis
i <- pseudoabsences(i, method = "bioclim", variables_selected = "vif") 
i
```

## Processing

### Modeling species relationship with variables

With the occurrences and predictors data put together, we can pass to the
modeling. As the name suggests, `caretSDM` uses the `caret` package underlying
its modeling procedure. For those who are not familiar, `caret` is the easiest
way to perform Machine Learning analysis in R. It works by setting a modeling
wrapper to pass multiple packages and can provide a lot of automation regarding
algorithms fine-tuning, data spliting, pre-processing methods and predictions.
These automated functions from `caret` can be altered using the `ctrl` argument
in `train_sdm` function. See `?caret::trainControl` for all options available.

We show here how to use a repeated crossvalidation method, which is defined
through `caret::trainControl`.

Note that, when you are using an algorithm for the first time, caret will ask
you to install the relevant packages to properly run the algorithm.

```{r train_sdm, results="hide", warning=FALSE, message=FALSE}
ctrl_sdm <- caret::trainControl(method = "repeatedcv", 
                                number = 4, # Number of folds
                                repeats = 1, # Number of complete sets of folds to compute
                                classProbs = TRUE,
                                returnResamp = "all",
                                summaryFunction = summary_sdm,
                                savePredictions = "all")

i <- train_sdm(i, 
               algo = c("mda", "naive_bayes", "kknn"), 
               crtl = ctrl_sdm, 
               variables_selected = "vif")
```

```{r train_sdm2}
i
```

```{r val_metrics}
get_validation_metrics(i)
write_validation_metrics(i, path = "results")
```

## Post-Processing

### Predicting species distribution in given scenarios

Now that we have our models, we can make predictions in new scenarios. The
function `predict_sdm` incorporates also the prediction of ensembles
(`ensembles=TRUE` is standard).

```{r predict_sdm}
i <- predict_sdm(i, th = 0.9)
i
```

```{r}
write_ensembles(i, path = "results/ensembles", ext = ".tif")
write_ensembles(i, path = "results/ensembles", ext = ".csv", centroid = TRUE)
```

```{r araucaria}
# Plot GCM ensemble:
plot_predictions(i, spp_name = "Araucaria_angustifolia")
```

```{r ilex}
plot_predictions(i, spp_name = "Ilex_paraguariensis")
```

```{r myrceugenia}
plot_predictions(i, spp_name = "Myrceugenia_euosma")
```
