---
title: "10. Ablation Analysis of Hyperparameter Tuning Length in caretSDM"
author: "[author name]"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: '3'
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
vignette: "%\\VignetteIndexEntry{10. Ablation Analysis of Hyperparameter Tuning Length in caretSDM} %\\VignetteEncoding{UTF-8}
  %\\VignetteEngine{knitr::knitr}\n"
editor_options:
  markdown:
    wrap: 100
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 5
)
```

------------------------------------------------------------------------

## Overview

Hyperparameter tuning is a critical step in machine-learning-based Species Distribution Modelling
(SDM). In `caretSDM`, tuning is controlled by the `tuneLength` argument passed to `train_sdm()`,
which determines how many candidate values are explored for each hyperparameter of every algorithm.
Larger values lead to a finer-grained search over the parameter grid, potentially yielding better
cross-validated performance --- but at the cost of longer computation times.

This document conducts an **ablation analysis** of `tuneLength` on the modelling of
*Araucaria angustifolia* (Brazilian pine), a critically endangered conifer endemic to the Atlantic
Forest and Southern Cone. Specifically, we ask:

> *By how much (in percentage) does increasing `tuneLength` improve model performance relative to
> the minimal single-value grid?*

The values evaluated are **`tuneLength` $\in$ {1, 3, 5, 10}**. All other modelling decisions
(algorithms, cross-validation scheme, pseudoabsence strategy, predictor selection) are held
constant so that observed differences in validation metrics are attributable solely to the tuning
budget.

------------------------------------------------------------------------

## 1 · Setup

### 1.1 Load packages

```{r packages}
library(caretSDM)
library(ggplot2)
library(dplyr)
```

### 1.2 Load built-in data

`caretSDM` ships with occurrence records and bioclimatic predictors for *A. angustifolia*. We load
them here and perform the standard preprocessing steps (data cleaning, multicollinearity
filtering, and pseudoabsence generation) **once**, before the ablation loop, since those steps are
independent of `tuneLength`.

```{r occ}
# Occurrence records: a data.frame with columns lon, lat
head(occ)

# Build occurrences_sdm object
oc <- occurrences_sdm(occ, occ_crs = 6933)
```

```{r bioc}
# Bioclimatic data covering the species' range retrieved using the following code:
WorldClim_data("current", variable = "bioc", resolution = 10)

# Import bioclimatic data
bioc <- raster::stack(list.files("current", pattern = ".tif", full.names = T)[c(1, 14, 4)])
names(bioc) <- c("bio1", "bio4", "bio12")

# Build sdm_area object
sa <- sdm_area(bioc,
  output_crs = 4326,
  crop_by = buffer_sdm(occ, size = 500000, occ_crs = 6933)
) |>
  add_scenarios()
```

### 1.3 Shared preprocessing

Data cleaning and pseudoabsence generation are performed once and reused across all tuning runs.
This isolates the effect of `tuneLength` from stochastic variation introduced by different
pseudoabsence sets.

```{r preprocessing}
set.seed(1)

# Pre-process: clean data, generate pseudoabsences
preprocessed <- input_sdm(oc, sa) |>
  data_clean() |>
  pseudoabsences(method = "random", n_set = 3)
```

------------------------------------------------------------------------

## 2 · Ablation workflow

A helper function wraps the training and prediction steps, accepting `tuneLength` as its only
varying argument. The cross-validation scheme uses 4-fold repeated cross-validation (4 repeats)
to ensure stable metric estimates.

```{r workflow-function}
run_tuning <- function(tune_length) {
  set.seed(1)
  preprocessed |>
    train_sdm(
      algo = c("naive_bayes", "kknn", "svmRadial", "nnet", "rf", "fda", "glm"),
      tuneLength = tune_length,
      ctrl = caret::trainControl(
        method = "repeatedcv",
        number = 4,
        repeats = 4,
        classProbs = TRUE,
        returnResamp = "all",
        summaryFunction = summary_sdm,
        savePredictions = "all"
      )
    ) |>
    predict_sdm(th = 0.9) |>
    ensemble_sdm()
}
```

------------------------------------------------------------------------

## 3 · Run the ablation

Each call uses an identical preprocessed object and identical cross-validation scheme; only
`tuneLength` varies. Running these sequentially can take several minutes --- this mirrors a typical
hyperparameter search cost analysis.

```{r run-ablation, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
tune_lengths <- c(1, 3, 5, 10)

sdm_list <- lapply(tune_lengths, run_tuning)
names(sdm_list) <- paste0("tuneLength_", tune_lengths)
```

------------------------------------------------------------------------

## 4 · Extract and compare performance metrics

### 4.1 Per-fold ROC/AUC (all resamples)

Raw fold-level metrics allow us to inspect the full distribution of ROC/AUC values for each
tuning budget, not just the mean.

```{r extract-metrics}
# Helper: pull all per-fold metrics for one SDM run
extract_metrics <- function(sdm_obj, tune_length) {
  get_validation_metrics(sdm_obj)[[1]] |>
    as.data.frame() |>
    mutate(tuneLength = tune_length) |>
    select(tuneLength, algo, ROC, ROCSD)
}

perf_all <- mapply(
  extract_metrics,
  sdm_obj = sdm_list,
  tune_length = tune_lengths,
  SIMPLIFY = FALSE
) |>
  bind_rows() |>
  mutate(tuneLength = factor(tuneLength, levels = tune_lengths))
```

### 4.2 Mean ROC/AUC per algorithm × tuneLength

```{r mean-metrics}
# Helper: pull mean metrics for one SDM run
extract_mean_metrics <- function(sdm_obj, tune_length) {
  mean_validation_metrics(sdm_obj)[[1]] |>
    as.data.frame() |>
    mutate(tuneLength = tune_length) |>
    select(tuneLength, algo, ROC, ROCSD)
}

mean_perf <- mapply(
  extract_mean_metrics,
  sdm_obj = sdm_list,
  tune_length = tune_lengths,
  SIMPLIFY = FALSE
) |>
  bind_rows() |>
  mutate(tuneLength = factor(tuneLength, levels = tune_lengths))

mean_perf
```

------------------------------------------------------------------------

## 5 · Percentage improvement over baseline

The baseline is `tuneLength = 1` --- the minimal single-value grid, equivalent to using default
algorithm parameters with no tuning. For each algorithm and each candidate `tuneLength`, we
compute the relative gain in mean ROC/AUC:

$$
\Delta_{\text{ROC}} = \frac{\overline{\text{ROC}}_{\texttt{tuneLength}=k} -
\overline{\text{ROC}}_{\texttt{tuneLength}=1}}
{\overline{\text{ROC}}_{\texttt{tuneLength}=1}} \times 100\%
$$

```{r pct-improvement}
# Baseline: tuneLength = 1
baseline <- mean_perf |>
  filter(tuneLength == 1) |>
  select(algo, ROC_base = ROC)

improvement <- mean_perf |>
  left_join(baseline, by = "algo") |>
  mutate(
    pct_improvement = (ROC - ROC_base) / ROC_base * 100,
    tuneLength = factor(tuneLength, levels = tune_lengths)
  )

# Overall mean improvement (averaged across algorithms) per tuneLength
overall_improvement <- improvement |>
  group_by(tuneLength) |>
  summarise(
    mean_ROC = mean(ROC),
    sd_ROC = sd(ROC),
    mean_pct_improvement = mean(pct_improvement),
    .groups = "drop"
  )

overall_improvement
```

------------------------------------------------------------------------

## 6 · Visualisations

### 6.1 ROC/AUC distribution across folds by algorithm and tuneLength

```{r boxplot, fig.height=5, fig.width=6}
ggplot(perf_all, aes(x = algo, y = ROC, fill = tuneLength)) +
  geom_boxplot(alpha = 0.75, outlier.size = 1.0) +
  geom_hline(yintercept = 0.9, linetype = "dashed", colour = "grey40") +
  ylim(0.5, 1) +
  scale_fill_manual(
    values = c(
      "1"  = "#4DA6C8",
      "3"  = "#6DBF67",
      "5"  = "#E07B54",
      "10" = "#9B59B6"
    ),
    labels = paste0("tuneLength = ", tune_lengths)
  ) +
  labs(
    title = expression(italic("Araucaria angustifolia") ~
      "- ROC/AUC by algorithm and tuning budget"),
    subtitle = "Dashed line: ROC/AUC = 0.9 (commonly used good-model threshold)",
    x = "Algorithm",
    y = "Validation Metric (ROC/AUC)",
    fill = "Tuning budget"
  ) +
  theme_bw(base_size = 10) +
  theme(legend.position = "bottom")
```

### 6.2 Percentage improvement over baseline (tuneLength = 1)

```{r pct-plot, fig.height=5, fig.width=6}
# Exclude tuneLength = 1 (baseline, improvement = 0%)
improvement_nonbaseline <- improvement |>
  filter(tuneLength != 1)

ggplot(improvement_nonbaseline, aes(x = algo, y = pct_improvement, fill = tuneLength)) +
  geom_col(position = position_dodge(width = 0.75), alpha = 0.85, width = 0.65) +
  geom_hline(yintercept = 0, colour = "grey30", linewidth = 0.4) +
  scale_fill_manual(
    values = c(
      "3"  = "#6DBF67",
      "5"  = "#E07B54",
      "10" = "#9B59B6"
    ),
    labels = paste0("tuneLength = ", c(3, 5, 10))
  ) +
  labs(
    title = expression(italic("Araucaria angustifolia") ~
      "-- % ROC/AUC gain vs. tuneLength = 1"),
    subtitle = "Positive values indicate improvement over the no-tuning baseline",
    x = "Algorithm",
    y = "Relative ROC/AUC improvement (%)",
    fill = "Tuning budget"
  ) +
  theme_bw(base_size = 10) +
  theme(legend.position = "bottom")
```

### 6.3 Mean ROC/AUC and percentage improvement across tuneLength values

```{r overall-trend-plot, fig.height=5, fig.width=6}
ggplot(overall_improvement, aes(x = as.integer(as.character(tuneLength)))) +
  geom_ribbon(
    aes(
      ymin = mean_ROC - sd_ROC,
      ymax = mean_ROC + sd_ROC
    ),
    fill = "#4DA6C8", alpha = 0.25
  ) +
  geom_line(aes(y = mean_ROC), colour = "#4DA6C8", linewidth = 1.1) +
  geom_point(aes(y = mean_ROC), colour = "#4DA6C8", size = 3) +
  geom_text(
    aes(
      y = mean_ROC,
      label = ifelse(
        tuneLength == 1,
        "baseline",
        paste0("+", round(mean_pct_improvement, 2), "%")
      )
    ),
    vjust = -1.2, size = 3.5
  ) +
  scale_x_continuous(breaks = tune_lengths) +
  labs(
    title = expression(italic("Araucaria angustifolia") ~
      "-- mean ROC/AUC across algorithms vs. tuneLength"),
    subtitle = "Ribbon: ±1 SD across algorithms. Labels: % gain over tuneLength = 1.",
    x = "tuneLength",
    y = "Mean ROC/AUC"
  ) +
  theme_bw(base_size = 10)
```

### 6.4 Heatmap of mean ROC/AUC per algorithm × tuneLength

```{r heatmap, fig.height=5, fig.width=6}
ggplot(mean_perf, aes(x = tuneLength, y = algo, fill = ROC)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = round(ROC, 3)), size = 3.2) +
  scale_fill_gradient2(
    low      = "#154360",
    mid      = "#1ABC9C",
    high     = "#FFC300",
    midpoint = 0.90,
    limits   = c(0.80, 1.00),
    name     = "Mean ROC/AUC"
  ) +
  labs(
    title = expression(italic("Araucaria angustifolia") ~
      "-- mean ROC/AUC heatmap"),
    x = "tuneLength",
    y = "Algorithm"
  ) +
  theme_minimal(base_size = 10)
```

------------------------------------------------------------------------

## 7 · Suitability maps across tuneLength values

Ensemble suitability maps allow us to inspect whether larger tuning budgets produce qualitatively
different spatial predictions, beyond differences captured by cross-validated metrics alone.

```{r maps-tl1, fig.height=5, fig.width=6}
plot_ensembles(sdm_list[["tuneLength_1"]])
```

```{r maps-tl3, fig.height=5, fig.width=6}
plot_ensembles(sdm_list[["tuneLength_3"]])
```

```{r maps-tl5, fig.height=5, fig.width=6}
plot_ensembles(sdm_list[["tuneLength_5"]])
```

```{r maps-tl10, fig.height=5, fig.width=6}
plot_ensembles(sdm_list[["tuneLength_10"]])
```

------------------------------------------------------------------------

## 8 · Quantitative comparison of spatial predictions

### 8.1 Pearson correlation matrix between ensemble predictions

```{r spatial-correlation}
pred_vals <- data.frame(
  tl1  = get_ensembles(sdm_list[["tuneLength_1"]])[1, 1][[1]][, "average"],
  tl3  = get_ensembles(sdm_list[["tuneLength_3"]])[1, 1][[1]][, "average"],
  tl5  = get_ensembles(sdm_list[["tuneLength_5"]])[1, 1][[1]][, "average"],
  tl10 = get_ensembles(sdm_list[["tuneLength_10"]])[1, 1][[1]][, "average"]
)

cor_mat <- cor(pred_vals, method = "pearson") |>
  as.table() |>
  as.data.frame()
```

```{r corrplot, fig.height=5, fig.width=6}
ggplot(cor_mat, aes(Var1, Var2, fill = Freq)) +
  geom_tile(colour = "black") +
  geom_text(aes(label = round(Freq, 3))) +
  scale_fill_gradient2(
    low      = "#154360",
    mid      = "#1ABC9C",
    high     = "#FFC300",
    midpoint = 0.975,
    limits   = c(0.95, 1.00)
  ) +
  coord_fixed() +
  scale_x_discrete(labels = paste0("tl=", tune_lengths)) +
  scale_y_discrete(labels = paste0("tl=", tune_lengths)) +
  theme_minimal() +
  ylab("") +
  xlab("")
```

High Pearson correlations between prediction surfaces indicate that the spatial structure of the
ensemble maps is largely preserved across tuning budgets. Conversely, lower correlations flag
regions where the tuning budget drives meaningful differences in predicted suitability.

------------------------------------------------------------------------

## 9 · Summary table

```{r summary-table}
summary_tbl <- improvement |>
  group_by(tuneLength) |>
  summarise(
    mean_ROC = round(mean(ROC), 4),
    sd_ROC = round(sd(ROC), 4),
    mean_pct_improvement = round(mean(pct_improvement), 2),
    max_pct_improvement = round(max(pct_improvement), 2),
    n_algo_improved = sum(pct_improvement > 0),
    .groups = "drop"
  ) |>
  rename(
    `tuneLength`                   = tuneLength,
    `Mean ROC/AUC`                 = mean_ROC,
    `SD ROC/AUC`                   = sd_ROC,
    `Mean % improvement`           = mean_pct_improvement,
    `Max % improvement (any algo)` = max_pct_improvement,
    `# algos improved`             = n_algo_improved
  )

summary_tbl
```

------------------------------------------------------------------------

## 10 · Discussion

**Effect of `tuneLength` on cross-validated performance.** Increasing the tuning budget generally
improves ROC/AUC, particularly for algorithms with high-dimensional hyperparameter spaces such as
`svmRadial`, `nnet`, and `kknn`. Simpler algorithms (e.g. `glm`, `fda`) benefit less because
their parameter grids are inherently small and may be exhausted at `tuneLength = 1`.

**Diminishing returns.** The percentage gain between `tuneLength = 1` and `tuneLength = 3` is
typically larger than the gain between `tuneLength = 5` and `tuneLength = 10`. This is expected:
the first few grid points of a coarse search tend to identify the most informative region of the
hyperparameter space, while finer refinements yield smaller marginal improvements. The
`overall_improvement` table and the trend plot in Section 6.3 make this pattern explicit.

**Spatial stability.** The high Pearson correlations between ensemble predictions across tuning
budgets (Section 8.1) suggest that, despite differences in cross-validated metrics, the spatial
structure of the final suitability maps is largely robust to the tuning budget. This is reassuring
for practitioners who prioritise map agreement over marginal metric gains.

**Practical guidance.** Based on these results, a `tuneLength` of 3--5 represents a good
compromise between model quality and computational cost for typical SDM workflows with
`caretSDM`. A `tuneLength = 1` (no tuning) is suitable for exploratory analyses or rapid
prototyping, whereas `tuneLength = 10` may be warranted for final, publication-quality models
where marginal performance gains matter.

------------------------------------------------------------------------

## Session information

```{r session-info}
sessionInfo()
```
