---
title: "Modeling Rare Species using Ensemble of Small Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modeling Rare Species using Ensemble of Small Models}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 80
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

Species Distribution Modeling (SDM) is a method majorly used to solve Wallacean shortfalls, i.e. it
aims to uncover the distribution of species. This workflow is very well established for species with
well studied and well collected species, which we have great availability of records. However, most
species are rare species, which even if well sampled would not have a lot of data to work with. In
this sense, modelers have developed methods to overcome the lack of data in SDMs. The most
widespread is the Ensemble of Small Models (ESMs), which builds SDMs using bivariate models and then
ensembles them into one result.

In caretSDM, ESMs area available through a simple function called use_esm. This function should be
used before training the model to adjust the settings properly. A compact code for ESM is the
following:

```{r esm_sdm}
library(caretSDM)
start_time <- Sys.time()
set.seed(1)

# Build sdm_area object
sa <- sdm_area(parana, 
               cell_size = 25000, 
               crs = 6933, 
               gdal = T) |> 
  add_predictors(bioc) |> 
  add_scenarios() |>
  select_predictors(c("bio1", "bio4", "bio12"))

# Build occurrences_sdm object
oc <- occurrences_sdm(occ, crs = 6933) |> 
  join_area(sa)

# Merge sdm_area and occurrences_sdm and perform pre-processing, processing and projecting.
i_esm <- input_sdm(oc, sa) |> 
  data_clean() |> 
  pseudoabsences(method = "bioclim") |> 
  use_esm(n_records = 999) |>
  train_sdm(algo = c("naive_bayes", "kknn"), 
            crtl = caret::trainControl(method = "repeatedcv", 
                                       number = 4, 
                                       repeats = 1, 
                                       classProbs = TRUE, 
                                       returnResamp = "all", 
                                       summaryFunction = summary_sdm, 
                                       savePredictions = "all")) |> 
  predict_sdm(th = 0.9) |> 
  suppressWarnings()
```

Note that we are setting `use_esm(n_records = 999)`, which means that we want every species with
number of records lower than 999. The standard value is 20, but the user can also specifically
inform which species should the ESM framework be applied to, by using the `spp` argument (e.g.:
`use_esm(spp = "Araucaria angustifolia")`).

## Comparing projections with standard SDM workflow

Now let's compare the outputs of ESM and standard SDMs. First, we need to create the standard SDM.
We will do that by just removing the `use_esm` function from previous chunk.

```{r standard_sdm}
i_sdm <- input_sdm(oc, sa) |> 
  data_clean() |> 
  pseudoabsences(method = "bioclim") |> 
  train_sdm(algo = c("naive_bayes", "kknn"), 
            crtl = caret::trainControl(method = "repeatedcv", 
                                       number = 4, 
                                       repeats = 1, 
                                       classProbs = TRUE, 
                                       returnResamp = "all", 
                                       summaryFunction = summary_sdm, 
                                       savePredictions = "all")) |> 
  predict_sdm(th = 0.9) |> 
  suppressWarnings()
```

Now let's compare both results:

```{r plot_esm}
# Plotting ESM result for current scenario
plot_predictions(i_esm,
                 spp_name = NULL,
                 scenario = "current",
                 id = NULL,
                 ensemble = TRUE,
                 ensemble_type = "mean_occ_prob")
```

```{r plot_standard_sdm}
# Plotting standard result for current scenario
plot_predictions(i_sdm,
                 spp_name = NULL,
                 scenario = "current",
                 id = NULL,
                 ensemble = TRUE,
                 ensemble_type = "mean_occ_prob")
```

## Comparing response curves

Another way to compare outputs is by plotting response curves. We can see that, when using ESMs, a
lot more models are built. This is because each model from the standard SDM is divided in multiple
bivariate models. However, a good level of concordance is achieved, when comparing to ESMs with
standard SDMs, as shown bellow.

```{r pdp_sdm_esm}
pdp_sdm(i_esm)
```

```{r pdp_sdm_standard}
pdp_sdm(i_sdm)
```

## Conclusion

This vignette demonstrates how to build Ensemble of Small Models using `caretSDM`. With a simple
function the user can automatically control which species should pass to ESM by setting a number of
records threshold or nominally indicate which species it wants to apply bivariate models.

```{r rmd_time}
end_time <- Sys.time()
end_time - start_time
```
